---
title: "Asymetric Brownian Motion (ABM)"
author: "Théo Gaboriau"
date: "27 11 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```



## ABM conditions to match  


This presents the conditions that have to be matched for the inheritance of traits variance (range) during speciation.
Lets call $V_a$ the variance of the ancestor at $t = T_s$ and $V_1, V_2$ the variance of its descendants at $t = T_s$.

```{r}
plot.new()
plot.window(xlim = c(1, 10), ylim = c(1, 10))
segments(c(2,5.1,5.1), c(5,5.2,4.8), c(4.9, 8, 8), c(5, 8.1, 1.9))
points(c(4.9,5.1,5.1),c(5,5.2,4.8),pch = 16,cex = 1)
text(c(4.7,5.1,5.1),c(5,5.2,4.8),labels = c(expression(V[a]), expression(V[1]), expression(V[2])),pos = c(3,3,1))
axis(1, at = c(2,5,8), labels = c(expression(T[a]), expression(T[s]), 0))

```


First, each variance has to be strictly positive:

**condition 1:** 

$$V_a > 0$$

$$V_1 > 0$$

$$V_2 > 0$$


Second, the ancestral variance has to be conserved:

**condition 2:**

$$V_a ≤ V_1 + V_2 ≤ 2V_a$$


Ideally, we should allow the parametrization to meet a simple Brownian motion such as:

$$V_a = V_1 = V_2$$

## Latest try


We consider a parameter controling $V_1$:

$$\rho = \frac{V_1}{V_a} \mbox{ with } 0 < \rho ≤ 1$$


We consider another parameter controling the amount of variance inherited by both $V_1$ and $V_2$:

$$\delta = \frac{V_2 + V_1 - V_a}{V_1}\mbox{ with }0 < \delta ≤ 1$$  


That way we have:


$$V_1=V_a \rho$$  


$$V_2=V_a(1+\rho(\delta - 1))$$  


That parametrization meets both **condition 1** and **condition 2**. We have a simple BM ($V_a = V_1 = V_2$) with $\rho = 1$ and 
$\delta = 1$  

With the log-transformation we simply have:

$$log(V_1) = log(V_a) + log(\rho)$$
$$log(V_2) = log(V_a) + log(1+\rho(\delta-1))$$

These equations are valid since \rho and \delta > 0

### example with $V_a = 5$

```{r}
v1 <- function(rho, delta, Va){
  return(Va*(rho))
}

v2 <- function(rho, delta, Va){
  return(Va*((1-rho)+delta*rho))
}

colvec <- colorRampPalette(rev(c("#d53e4f", "#f46d43", "#fdae61", "#fee08b", "#ffffbf", "#e6f598", "#abdda4",
                              "#88ddaa", "#66c2a5", "#37b69b", "#3288bd", "#71acbc")))(101)
rhos <- seq(1e-3, 1, length = 100)
deltas <- seq(1e-3, 1, length = 100)
Va <- 5
out_v1 <- matrix(NA, 100, 100)
dimnames(out_v1) <- list(rhos, deltas)
out_v2 <- matrix(NA, 100, 100)
dimnames(out_v2) <- list(rhos, deltas)

for(i in 1:100){
rho <- rhos[i]
out_v1[i,] <- v1(rho, deltas, Va)
out_v2[i,] <- v2(rho, deltas, Va)
}

layout(t(1:2))
par(mar = c(10,5,5,1), las = 1)
plot(rep(rhos, times = 100), rep(deltas, each = 100), pch = 15, col = colvec[round(out_v1/max(c(out_v1, out_v2))*100)+1],
     main = expression(paste("Effect of ", rho, " and ", delta, " on ", v[1])), xlab = expression(rho), ylab = expression(delta)) 
hist(out_v1, main = expression(V[1]),breaks = seq(min(out_v1),max(out_v1), length = 100), col = colvec, border = NA, xlab = expression(V[1]))


plot(rep(rhos, times = 100), rep(deltas, each = 100), pch = 15, col = colvec[round(out_v2/max(c(out_v1, out_v2))*100)+1],
     main = expression(paste("Effect of ", rho, " and ", delta, " on ", v[2])), xlab = expression(rho), ylab = expression(delta))
hist(out_v2, main = expression(V[2]),breaks = seq(min(out_v2),max(out_v2), length = 100), col = colvec, border = NA, xlab = expression(V[2]))


plot(rep(rhos, times = 100), rep(deltas, each = 100), pch = 15, col = colvec[round((out_v1+out_v2)/max((out_v1+ out_v2))*100)+1],
     main = expression(paste("Effect of ", rho, " and ", delta, " on ", v[1] + v[2])), xlab = expression(rho), ylab = expression(delta)) 
hist(out_v1 + out_v2, main = expression(V[1]+V[2]), breaks = seq(min(out_v1 + out_v2),max(out_v1 + out_v2), length = 100), col = colvec, border = NA, xlab = expression(V[1] + V[2]))




```


The parameters here don't have a real meaning but we can interpret:  
 - the asymetry between the inherited variance of descendants 
 
 
 $$\eta =\frac{min(V_1,V_2)}{max(V_1,V_2)}=\begin{cases} \frac{\rho}{\rho\delta-\rho+1}&\mbox{if }V_1 < V_2\\1& \mbox{if } V_1 = V_2\\\delta - 1 + \frac{1}{\rho} & \mbox{if }V_1 > V_2 \end{cases}\mbox{ with }0≤\eta≤1$$
 - the variance partitioning between descendants  
 
 
 $$\omega = 1-(\frac{V_1+V_2-V_a}{V_a})= 1-\rho\delta \mbox{ with } 0≤\omega≤1$$
 

```{r, fig.height=5, fig.width=7,}
eta <- function(rho, delta){
  eta <- rho/(rho*delta-rho+1)
  eta[eta>1] <- 1/eta[eta>1]
  return(eta)
}

omega <- function(rho, delta){
  return(1-rho*delta)
}

colvec <- colorRampPalette(rev(c("#d53e4f", "#f46d43", "#fdae61", "#fee08b", "#ffffbf", "#e6f598", "#abdda4",
                              "#88ddaa", "#66c2a5", "#37b69b", "#3288bd", "#71acbc")))(101)
rhos <- seq(1e-3, 1, length = 100)
deltas <- seq(1e-3, 1, length = 100)
Va <- 5
out_eta <- matrix(NA, 100, 100)
dimnames(out_eta) <- list(rhos, deltas)
out_omega <- matrix(NA, 100, 100)
dimnames(out_omega) <- list(rhos, deltas)

for(i in 1:100){
rho <- rhos[i]
out_eta[i,] <- eta(rho, deltas)
out_omega[i,] <- omega(rho, deltas)
}

layout(t(1:2))
par(mar = c(10,5,5,1), las = 1)
plot(rep(rhos, times = 100), rep(deltas, each = 100), pch = 15, col = colvec[round(out_eta*100)+1],
     main = expression(paste("Effect of ", rho, " and ", delta, " on ", eta)), xlab = expression(rho), ylab = expression(delta)) 
hist(out_eta, main = expression(eta),breaks = seq(min(out_eta),max(out_eta), length = 100), col = colvec, border = NA, xlab = expression(eta))


plot(rep(rhos, times = 100), rep(deltas, each = 100), pch = 15, col = colvec[round(out_omega*100)+1],
     main = expression(paste("Effect of ", rho, " and ", delta, " on ", omega)), xlab = expression(rho), ylab = expression(delta))
hist(out_omega, main = expression(omega),breaks = seq(min(out_omega),max(out_omega), length = 100), col = colvec, border = NA, xlab = expression(omega))

```
 

## BM simulations 3 sp and $\sigma^2 = 0.1$  

```{r fig.height=5, fig.width=7, message=FALSE, warning=FALSE, include=FALSE}
# asymetric inheritance of x during speciation
# poisson process for speciation
sim_ABM <- function(theta0 = 0, sigma2 = 1, rho = 1, delta = 1, lambda = 5, nspmax = 100, ngen = 100, bounds = c(0, Inf)){
  
  ## initial conditions
  x <- matrix(NA, nrow = nspmax, ncol = ngen+1)
  x[1,1] <- theta0 
  t.sp <- rep(NA,nspmax)
  t.sp[1] <- rpois(1, lambda)
  i.sp <- rep(NA, nspmax)
  i.sp[1] <- 0
  spe <- rep(F, nspmax)
  
  ## function to respect bounds
  reflect <- function(yy, bounds) {
    if(!is.na(yy)){
      while (yy < bounds[1] || yy > bounds[2]) {
        if (yy < bounds[1]) 
          yy <- 2 * bounds[1] - yy
        if (yy > bounds[2]) 
          yy <- 2 * bounds[2] - yy
      }
    }
    return(yy)
  }
  
  ## simulation starts
  for (i in 2:(ngen+1)){
    nsp <- sum(!is.na(i.sp))
    
    # test for speciation
    spe[!is.na(i.sp)] <- i.sp[!is.na(i.sp)] == t.sp[!is.na(i.sp)]
    i.sp <- i.sp + 1
    
    if (any(spe) & nsp + sum(spe) <= nspmax){
      # asymetric inheritance of trait
      x[,i] <- x[,i-1]
      x[spe,i] <- rho * x[spe, i-1]
      x[nsp+1:sum(spe),i] <- (1+rho*(delta-1)) * x[spe, i-1]
      x[nsp+1:sum(spe),i-1] <- x[spe, i-1]
      
      # trait evolution under brownian motion
      x[,i] <- rnorm(nrow(x), x[,i], sigma2)
      x[,i] <- sapply(x[,i], reflect, bounds)
      
      # fill vector for speciation process
      i.sp[spe] <- 0
      t.sp[spe] <- rpois(sum(spe),lambda)
      i.sp[nsp+1:sum(spe)] <- rep(0, sum(spe))
      t.sp[nsp+1:sum(spe)] <- rpois(sum(spe),lambda)
      
    } else {
      
      # trait evolution under brownian motion
      x[,i] <- rnorm(nrow(x), x[,i-1], sigma2)
      x[,i] <- sapply(x[,i], reflect, bounds)
      
    }
  }
  return(x)
}


plot_ABM_sim <- function(sim, col, new = T, xlim, ylim){
  if(new){
    plot.new()
    plot.window(xlim = xlim, ylim = ylim)
    axis(2)
    mtext("trait variance", side = 2, line = 2)
    axis(1)
    mtext("generations", side = 1, line = 2)
  } 
  for(i in 1:nrow(sim)){
    lines(sim[i,], col = col)
  }
}


pars <- cbind(delta = c(0.1, 0.9), rho = c(0.1, 0.9))
comb <- rbind(c(1,1), c(1,2), c(2,1), c(2,2))
sim <- lapply(1:nrow(comb), function(i){
  x <- comb[i,]
  set.seed(300)
  sim_ABM(lambda = 4000, delta = pars[x[1],1], rho = pars[x[2],2], ngen = 1e4, nspmax = 3, sigma = 0.01, theta0 = 2)
})


```

```{r echo=FALSE, fig.height=7, fig.width=7, message=FALSE, warning=FALSE, results = 'hide'}
layout(cbind(c(1,2), c(3,4)))
par(mar = c(4,3,2,2)+1)
# delta = 0.1, rho = 0.1
plot_ABM_sim(sim[[1]], "#518eb6", T, xlim = c(0, ncol(sim[[1]])), ylim = c(0, max(sim[[1]], na.rm = T)))
mtext(expression(group("{",list(delta == 0.1, rho == 0.1), "}")), 3, line = 1)
# delta = 0.1, rho = 0.9
plot_ABM_sim(sim[[2]], "#518eb6", T, xlim = c(0, ncol(sim[[2]])), ylim = c(0, max(sim[[2]], na.rm = T)))
mtext(expression(group("{",list(delta == 0.1, rho == 0.9), "}")), 3, line = 1)
# delta = 0.9, rho = 0.1
plot_ABM_sim(sim[[3]], "#518eb6", T, xlim = c(0, ncol(sim[[3]])), ylim = c(0, max(sim[[3]], na.rm = T)))
mtext(expression(group("{",list(delta == 0.9, rho == 0.1), "}")), 3, line = 1)
# delta = 0.9, rho = 0.9
plot_ABM_sim(sim[[4]], "#518eb6", T, xlim = c(0, ncol(sim[[4]])), ylim = c(0, max(sim[[4]], na.rm = T)))
mtext(expression(group("{",list(delta == 0.9, rho == 0.9), "}")), 3, line = 1)

```


